---
title: "How to use RDS?"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How to use RDS?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  echo = TRUE,
  comment = "#>"
)
```

## Intro on RDS: Respondent-Driven Sampling

Hidden and hard to reach populations: No sampling frame​

Often want to remain hidden due to practices and behaviors that are illegal or stigmatized

Assumptions: 
Respondents know each other as members of the target population (and these are reciprocal)​

Respondents are linked by a network composed of a single component​

Sampling occurs with replacement (population from which a sample is gathered is infinitely large)​

Respondents can accurately report their personal network size (and this needs to be measured)​

Peer recruitment is a random selection from the recruiter’s network​

Each respondent recruits a single peer


* Sampling design:  Require respondents to choose from among their social circle rather than the researcher chooses.
* Seed Dependence: follow only a few links from each sampled
* Confidentiality:respondent-driven:respondents  distribute  uniquely  identified coupons. no names.
* Estimation: Several approaches
* Effective at obtaining large varied samples in many populations.
* Widely used: over 100 studies, in over 30 countries. Often HIV-risk populations.

Heckathorn, D.D.,“Respondent-driven sampling:  A new approach to the study ofhidden populations.”Social Problems, 1997.Salganik, M.J. and D.D. Heckathorn,“Sampling and estimation in hidden populationsusing respondent-driven sampling.”Sociological Methodology, 2004.


Ensure very long recruitment chains in RDS (consider coupon reduction techniques, 2 vs. ≥3 for populations with dense networks).​

Have fewer, well selected seeds; direct them on who to recruit. ​

Use diagnostics on already collected data to minimize bottlenecks in future rounds of data collection.​

Monitor recruitment daily​

Consider “narrowing” eligibility criteria to exclude problematic sub-populations.


## Set Up: load packages

```{r message=FALSE, warning=FALSE}
library(RDS)
library(ggplot2)
library(tidyverse)
```


## Import, Export and Reformat Data

### Import Data

If you are importing RDS data from a non-RDS file format (e.g., .xls or .csv), it must have certain variables defined, and those variables must be in specific formats, for it to be valid RDS data. They are:

 *  __Subject ID__: Each row of the data, i.e. each survey respondent, must be identified with a unique subject ID. This will mostly likely be an identification number (integer or double format) or an anonymous label (character format) associated with the respondent. There should not be any missing or NA values in the Subject ID column.
 
 * __Network Size__: Each respondent's self-reported network size (or degree), i.e. their number of associations (or alters) in the target population. Ideally, there are no missing or NA values in this column.
 
 * __Recruitment time__: An optional date or numeric variable indicating the time that each subject was recruited.
 
 * __Recruitment Information__: This tracks the path of recruitment in the sample and is either in Coupon Format or Recruiter ID Format. Upon loading data that is not already in RDS format, you will be prompted to convert the data to Coupon or Recruiter ID format. For more on this process, see RDSAnalystConvertToRDS. Note, you can leave the data un-converted if you do not want to use the RDS analysis features. 

### Convert Data to RDS Format

`as.rds.data.frame` RDS Object:		*.rdsobj, *.rdsat


Whether you automatically access the data conversion function upon uploading data or choose it from the Data menu, there are two possible RDS data formats:

#### Coupon format

In this format there are several columns in the data containing coupon numbers. One column is for each respondent's own coupon, and the others (e.g. coupon.1, coupon.2, etc.) list the coupons they may have distributed. Thus, the recruitment path is stored through the coupon numbers. The seeds generally have own coupon numbers of 0, NA or something similar. Upon conversion, columns identifying each respondent's recruiter ID, wave and seed are created if not already present.

#### Recruiter ID format

Rather than tracking recruitment paths through coupons, in this format a Recruiter ID column records the subject ID of the recruiter for each respondent. Seeds are identified by a shared recruiter ID such as 0. Upon conversion, columns identifying each respondent's wave and seed are created if not already present.
Example: Convert nyjazz data set to RDS coupon format

#### Common Data Problems & Solutions

 * Each subject must have a unique non-missing ID Symptom Error message in console when importing -> Solution 1.Edit data to insert missing ids 2.Remove extraneous rows 

 * Importing Coupon Format Problem The subject’s coupon must match exactly one of the coupon variables in the recruiter’s row. Symptom Incorrect recruitment tree plots Solution 1.Fix data entry errors in coupon variables. 2.Look for trailing and/or leading white space (spaces) in the coupon variables 

 * Importing Recruiter ID Format Problem The subject’s coupon must match exactly one of the coupon variables in the recruiter’s row. Symptom Error shown in console, or incorrect recruitment tree plots Solution 1.Fix data entry errors in the subject ID and recruiter ID variables. 2.Look for trailing and/or leading white space (spaces). 

 * Importing Recruiter ID Format Problem All seeds must have the same recruiter ID which can not be present in the Subject ID Symptom Error shown in console or incorrect recruitment tree plots Solution 1.Manually edit the recruiter id’s of the seeds. 2.Look to make sure the seeds’ recruiter id is not present in the subject id. 

 * Rounding error in IDS Problem Coupon or ID variables are numeric and some have 15 or more digits.  Symptom Recruitment structure incorrectly read in except when using .rdsat and .rdsobj formats. Solution 1.Edit variables to have non-numeric type (e.g. add a “c” to the beginning of all coupon variables). 2.Put data file in .rdsat format


Example on nyjazzdf 

```{r }
#mtcars[c("V8")] <- recode.variables(mtcars[c("cyl")] , "8 -> 1;else -> 0;") 
```


### Edit RDS Meta Data

An RDS data set's "Meta Data" stores information about the variable names associated with: 

 *  __Subject ID__: Uniquely identifies survey participants. It should not contain and NA values.
 *  __Recruiter ID__: should be a subset of subject ID values except for seeds who share a recruiter ID not used in Subject ID, e.g. 0. Also should not have NA values.
 *  __Network Size__: a.k.a Degree, is the number of connections in the network that each respondent reports.
 *  __Max # of Coupons__: The number of recruitment coupons distributed to each participant.
 *  __Population Size Estimate__: Mid values may be used as the default population size. 
 
 

### Recode Variables

The Recode Variables dialog allows to substitute values into existing variables, or create new variables based on current variables.

The following example uses the mtcars data set, which can be loaded by entering data(mtcars) into the console. 

```{r }
#mtcars[c("V8")] <- recode.variables(mtcars[c("cyl")] , "8 -> 1;else -> 0;") 
```




### Export (Save) Data



The Export Data dialog (under the File menu) allows storage of RDS data in several formats:

 *  __RDS Analyst (.rdsobj)__ - Stores a RDS data frame in the native RDS analyst format, and does not need to be converted if re-imported to RDS Analyst (see RDSAnalystImportRDSData)
 *  __Flat File (.robj, .csv)__ - Uses the dput function to write an ASCII text representation of a data frame to a file or connection (.robj).
 *  __ Netdraw (.dl, .vna)__ - Stores a RDS data frame's recruitment tree in NetDraw format
 *  __GraphViz (.gz)__ - Stores a RDS data frame's recruitment tree as a GraphViz file 



## Visualize and Summarize Data


### Plot the Recruitment Tree 

The Plot Recruitment Tree produces a publication quality graphics plot of the recruitment tree. As shown below, you can use variables from the data to set the node color, size and labels of the recruitment tree. 

Are the chains long? 
Do most of the subjects come from the same seed?
Visualize homophily.



### Plot Recruitment Diagnostics
The Recruitment Diagnostics creates up to five charts, providing a snapshot of the recruitment process:

 * Network Size by wave  
 * Recruits by wave  
 * Recruits by Seeds  
 * Recruits by subject  
 
Does network size change over the course of sampling? 
How many recruits are in each wave? 
How many recruits originate from each seed?

### Plot Basic Statistics

Continuous Variables 

 * Look at descriptive statistics (Sample > Descriptives)
 * Plot histograms (Plots > Quick > Histogram) 

Categorical Variables

* Look at Frequencies (Sample > Frequencies) 
* Make bar charts (Plots > Quick > bar)

Relationships

 * Crosstabs (Sample > Contingency Tables > chi-squared) 
 * Descriptives with strata (Sample DescriptivesL mean, standard variation)

## Analyze Data



### Population Frequency Estimates

This dialog computes a point estimate and confidence interval of population frequency (prevalence), for one or more categorical variables. 

The default method is Gile's SS estimator, where the confidence interval is computed using Gile's bootstrap method. This is a computationally-intensive procedure and can take a minute or longer to complete.

#### Convergence Plot

Convergence plots help to determine whether the final RDS estimate is biased by the initial convenience sample of seeds. If the estimates shown in the plots do not level off as the sample size increases then the current estimate is probably biased by the choice of seeds. It is recommended to create a convergence plot of all traits of interest during the research process and extend sampling if the estimate has not stabilized.

measures progression of enrolling subjects to determine when the proportion for a characteristic approaches and remains stable in relation to the adjusted estimate

The convergence plots below show a data set that has converged (left, using fauxmadrona sample data) and one that has not (right).

Note that although the plot on the left shows convergence, there is a chance that additional data would not follow this trend. The recruitment tree could be stuck in a sub-population of the population of interest. This is discussed further below.

#### Bottleneck Plot

Bottleneck plots are used to asses whether the population of interest contains distinct sub-communities that could bias the RDS estimate. If the sub-communities tend to recruit from within (i.e. exhibit homophily) then the choice of seeds may generate a biased sample. If the sub-communities are different with respect to the population characteristic of interest then the RDS prevalence estimate may also be biased.

Estimate precision is based on underlying network structure and traits within the structure.​

Assume that the structure makes up one complete social network component.​

Some structures have sub-populations not connected within the network structure.​

Especially problematic when traits differ between unconnected sub-populations in a sample.

An example of this scenario is described in "Diagnostics for Respondent-driven Sampling" (Gile, Johnston, Salganik 2012):

"Imagine a city with street-based sex workers and brothel-based sex workers where there are many social connections within these groups, but few connections between these groups. Further, imagine that brothel-based sex workers use condoms regularly, whereas street-based sex workers do not. This situation will be problematic for RDS because the network "bottleneck" between the two groups will prevent the sample from exploring the entire population and could lead to inaccurate estimates about both sex worker type (i.e., brothel-based vs. street-based) and condom usage."

Each line of a bottleneck plots tracks an estimate of a characteristic of an interest based on one of the seeds. If the estimates converge as the samples from each seed (their trees) grow, then there is no indication of a bottleneck along this characteristic.

Below are two bottleneck plots prepared from the fauxmadrona data set. (To load fauxmadrona, select "Example: fauxmadrona" from the Packages & Data menu). Both track the samples originating from each of the 10 seeds, as labeled by ID in the legend at right. (Note that the lines for each seed end at different places, reflecting the different sizes of the trees that originate from each seed.) The plot on the left tracks the estimate of the prevalence of disease in the population. While the estimates start at a range of values, they all converge toward about 20%, showing no evidence of divided sub-communities of infected and uninfected. Similarly, the plot at right tracks the samples along a completely randomized 0/1 indicator variable for the population. As expected, the prevalence estimates of this indicator converge toward .5, with some variation around that estimate.

In comparison, a plot with evidence of bottlenecks is shown below. The estimates stabilize at different places depending on where the seeds originate.


### Population Contingency Tables (Crosstabs)
The Population Crosstabs creates contingency tables for the target population based on inference from the sample. The tables show the estimated marginal and joint distribution of characteristics in the population. The column and row totals give point estimates (prevalence rates) for the selected variables -- in the example below, the row totals give the percentage of "blue" and "red" in the population based on RDS-II sampling weights. The tables can optionally be stratified by another variable in the data. Below they are stratified by wave.

Disclaimer: These tables do not show the uncertainty in the estimates. To get a sense of the uncertainty, use the Frequency Estimates dialog, which returns confidence intervals along with point estimates. The point estimates from that dialog should approximately match the ones here if the selected weight type and other settings match.

If you would like to view a contingency table for the sample data alone (no inference on the population) use the Contingency Tables dialog under the Sample menu
 

### Test Difference in Population Proportions / Prevalence

The Test Difference in Proportions is used to compare population frequency estimates. The table output of the population Frequency Estimates dialog is also produced here, but this dialog adds a comparison table.

For example, the following compares the frequency of classes of the X and Y variables of the "faux" data set.



### Compute Sampling Weights

he Compute Weights dialog under the Data menu to calculate sampling weights based on reported network size (degree). This dialog will store weights in your data set as a new variable (column) in the data set. Weights may be used for fitting a generalized linear model to the sample data. Several other dialogs in the software ask you to choose a weight type as part of their calculation, e.g. Frequency Estimates and Population Crosstabs. The methods available for calculating weights are described below. (In some dialogs, Arithmetic Mean is also listed as a weight type, which simply means that all members of the sample are weighted equally.)

 * __HCG (Homophily Configuration Graph estimator)__: Recommended when the sample is a significant fraction of the target population and recruitment time is known. An estimate of the population size is required to use this estimator.
 * __Gile's SS (Sequential Sampler)__: Recommended when the sample is a significant fraction of the target population. It is based on the inclusion probabilities of members of the sample, which are based on reported network sizes (how many people a respondent knows within the target population). An estimate of the population size is required to use this estimator.
 * __RDS-I: The Salganik-Heckathorn estimator__. RDS-I weights are calculated in reference to a particular categorical variable of interest (the Group Variable; it cannot be continuous). RDS-I calculations are based on the number of connections between "in-group" (e.g. infected) and "out-of-group" (e.g. uninfected) members of the sample (though more than two classes are possible). A Markov process is used to model population mixing, i.e. recruitment across groups, and generate an equilibrium estimate of group prevalence (e.g. infection rate).
 * __RDS-I (DS): Data-smoothed version of RDS-I__. It is assumed that the Markov process is reversible.
 * __RDS-II: The Volz-Heckathorn estimator__, a generalized Horvitz-Thomspon form. Like Gile's SS, it is based on inclusion probabilities for members of the sample, which are based on reported network sizes. It treats the sampling process as a random walk through the network of the target population. 

For Gile's SS and RDS-II, weights will be unique for each reported network size. For RDS-I and RDS-I (DS), weights will be unique for each class of the Group Variable.



### Recruitment Homophily & Population Homophily

Compute a homophily measure for the recruitment process, i.e. do respondents tend to recruit people like themselves? 

Take as an example homophily by disease status. In this case, it is the ratio of number of recruits that have the same disease status as their recruiter to the number we would expect by chance. The difference with the Population Homophily is that here we calculate for the recruitment chain rather than the whole population of social ties.

If the recruitment homophily on disease status is about 1, we see little effect of recruitment homophily.
 Values larger than one indicate more homophily. 

The following shows how to calculate recruitment homophily on disease status for the fauxmadrona data set. 

Tendency of like to recruit like. 
Lots of homophily -> High Variance 

Homophily near 1 means no homophily
p-values assume a simple random sample, so only use them as rough guides. 



Population Homophily is very similar to recruitment homophily, except here we estimate the homophily for the population as a whole, and the function only outputs that estimate (no other descriptives). The function demands an additional estimate of the population size, which can be calculated using the size package. (Note, the homophily estimate may not be very sensitive to the population size estimate.) For further description of homophily, see the recruitment homophily page.


### Compare Network Activity of Sub-populations: Differential Activity

The distribution of network size (a.k.a. degree) of members of the target population may depend on some characteristic(s) of the population. For example, if the characteristic of interest is the prevalence of a contagious disease, those with the disease may also have a higher average number of connections in the social network. The differential activity dialog compares the (weighted) average network size of two classes of the population, e.g. those with and without a disease.

The user inputs the choice of data set and binary variable(s) on which to compare members of the population, e.g. disease status.

The output is, for example:
"The mean degree of those with value 1 divided by the mean degree of those without is 0.6189263"

Note that "those with value 1" refers to the group that is ordered second alphabetically or numerically. If the variable has named states, e.g. "blue" and "red", the function will order alphabetically by default. If the states are integers, "those with value 1" would be those with the higher-number state.


### Test of trend

This function compute a hypothesis test of trend in prevalence based on a likelihood-ratio statistic. It takes a series of point estimates and their associated standard errors and computes the p-value for the test of a monotone decrease in the population prevalences (in sequence order). 

The p-value for a monotone increase is also reported. An optional plot of the estimates and the null distribution of the test statistics is provided. 
